---
title: "Flat"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Flat}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Constrained Smoothers

This package provides smooths constrained to have zero first derivative at one
or both of the leftmost and rightmost knots.  There are two forms: the "z"
smooths constrain the first derivative to be zero but the smooth is a cubic
beyond that; while the "f" smooths constrain the first derivative to be zero but
the smooth is linear beyond that.  Each has three forms, "zb", "zl", "zr" and
"fb", "fl", "fr" depending upon whether the derivative should be zero at both,
or the just leftmost or rightmost knots respectively.

This is a straightforward extension of Wood (2017), section 5.3.1.

```{r setup}
library(flat)
```

## Noisy Cosine

We illustrate the smooths with a noisy cosine.  In the first case the knots
contain the data, in the second the data extend beyond the knots.

### Test 1

Generate cosine data on the interval [0,1]
```{r}
library(mgcv)
set.seed(37)
n <- 500
d <- data.frame(x=seq(0,1,length=n),
                y=cos(pi*seq(0,1,length=n))+rnorm(n,0,0.01))
ks <- seq(0,1,length=10)
plot(y~x,data=d)
```

Fit a standard cubic spline
```{r}
fit <- gam(y ~ s(x,bs="cr"),
           knots = list(x=ks),
           family=gaussian(),data=d)
summary(fit)
fit$aic
plot(fit,ylim=c(-1.5,1.5))
```

Constrain the smooth to be flat at both ends
```{r}
fit <- gam(y ~ s(x,bs="fb"),
           knots = list(x=ks),
           family=gaussian(),data=d)
summary(fit)
fit$aic
plot(fit,ylim=c(-1.5,1.5))
```


Constrain the smooth to be flat at the leftmost knot
```{r}
fit <- gam(y ~ s(x,bs="fl"),
           knots = list(x=ks),
           family=gaussian(),data=d)
summary(fit)
fit$aic
plot(fit,ylim=c(-1.5,1.5))
```

Constrain the smooth to be flat at the rightmost knot
```{r}
fit <- gam(y ~ s(x,bs="fr"),
           knots = list(x=ks),
           family=gaussian(),data=d)
summary(fit)
fit$aic
plot(fit,ylim=c(-1.5,1.5))
```


### Test 2

Generate cosine data on the interval [-0.2,1.2]
```{r}
library(mgcv)
set.seed(37)
n <- 500
d <- data.frame(x=seq(-0.2,1.2,length=n),
                y=cos(pi*seq(-0.2,1.2,length=n))+rnorm(n,0,0.01))
ks <- seq(0,1,length=10)
plot(y~x,data=d)
```

Fit a standard cubic spline
```{r}
fit <- gam(y ~ s(x,bs="cr"),
           knots = list(x=ks),
           family=gaussian(),data=d)
summary(fit)
fit$aic
plot(fit,ylim=c(-1.5,1.5))
```

Constrain the smooth to be flat at both ends
```{r}
fit <- gam(y ~ s(x,bs="fb"),
           knots = list(x=ks),
           family=gaussian(),data=d)
summary(fit)
fit$aic
plot(fit,ylim=c(-1.5,1.5))
```

Constrain the smooth to have zero derivative at both ends
```{r}
fit <- gam(y ~ s(x,bs="zb"),
           knots = list(x=ks),
           family=gaussian(),data=d)
summary(fit)
fit$aic
plot(fit,ylim=c(-1.5,1.5))
```

Constrain the smooth to be flat at the leftmost end
```{r}
fit <- gam(y ~ s(x,bs="fl"),
           knots = list(x=ks),
           family=gaussian(),data=d)
summary(fit)
fit$aic
plot(fit,ylim=c(-1.5,1.5))
```

Constrain the smooth to have zero derivative at the leftmost end
```{r}
fit <- gam(y ~ s(x,bs="zl"),
           knots = list(x=ks),
           family=gaussian(),data=d)
summary(fit)
fit$aic
plot(fit,ylim=c(-1.5,1.5))
```


Constrain the smooth to be flat at the rightmost end
```{r}
fit <- gam(y ~ s(x,bs="fr"),
           knots = list(x=ks),
           family=gaussian(),data=d)
summary(fit)
fit$aic
plot(fit,ylim=c(-1.5,1.5))
```

Constrain the smooth to have zero derivative at the rightmost end
```{r}
fit <- gam(y ~ s(x,bs="zr"),
           knots = list(x=ks),
           family=gaussian(),data=d)
summary(fit)
fit$aic
plot(fit,ylim=c(-1.5,1.5))
```
